Wildcard: Spreadsheet-Driven Customization of Web Applications

### 2020-01-10 paper.tex
nice draft

In 2012, the travel site Airbnb removed the ability to sort listings by
price. Users could still filter by price range, but could no longer view
the cheapest listings first. Many users complained on online message
boards that the change seemed hostile to users. ``It's so
frustrating!..What is the logic behind not having this function?'' said
one user on the
\href{https://community.withairbnb.com/t5/Hosting/Sorting-listing-by-price/td-p/559404}{Airbnb
support forum}. Alas, the feature remains missing to this day.

This is a familiar situation in a world of web applications that are
frequently updated without user consent. Sometimes there is a browser
extension or user script that fixes an issue, and if the user is both
motivated and skilled they might even be able to implement their own
fix. But for most people, the only recourse is to complain to the
developers and pray that someone listens---or more likely, to simply
give up. While many have become used to this status quo, we see it as a
tremendous waste of the openness of the Web platform. Back in 1977, in
\emph{Personal Dynamic Media} \autocite{kay1977}, Alan Kay originally
envisioned personal computing as a medium that let a user ``mold and
channel its power to his own needs,'' but today software feels more like
concrete than clay.

In this paper, we present \emph{spreadsheet-driven customization}, a
technique for making software customization more accessible to end
users. The core idea is to show the data inside the application in a
spreadsheet which maintains a live connection to the original UI. When
the user manipulates the spreadsheet, the UI is instantly modified, and
vice versa. We built a research prototype called Wildcard that
implements spreadsheet-driven customization as a web browser extension,
showing a spreadsheet panel in the context of existing sites.

Spreadsheets have proven to be a widely successful computing platform
for non-programmers. They provide a low barrier to entry for beginners,
while allowing experts to perform complex computations. The 2D grid is
simple to grasp, but general enough to represent many types of data.
Prior work \autocite{mccutchen2016,benson2014,chang2014} has used these
benefits to enable end users to easily create web applications that use
spreadsheets as a backing data layer.

Spreadsheet-driven customization applies these exact same benefits in a
different context: customizing existing software, rather than building
new software from scratch. We inherit the low floor and high ceiling of
spreadsheets: small tweaks like sorting a list of data in an app can be
done with a single click, while more complex customizations, like
joining in related data from a web API, are supported by a rich formula
system. The tabular data format is versatile, supporting data from a
wide range of applications. Spreadsheet-driven customization does not
require that the application is actually backed by a spreadsheet---it
merely presents a model of the internal state of the application, which
can be exposed by the application itself or derived from the user
interface by third parties. This means that spreadsheet-driven
customization works with existing applications that people already use.
It also allows for exposing a common data abstraction across
applications, enabling users to have a consistent mechanism for
manipulating data from all of their applications.

Prior work \autocite{huynh2006} has shown that the process of extracting
structured data from existing user interfaces can be confusing and
unfamiliar for end users. In order to make the customization experience
as straightforward as possible, we envision a tiered architecture that
hides this complexity from end users. Wildcard provides a mechanism for
programmers to write \emph{site adapters}, which use web scraping
techniques to extract structured data from the DOM and AJAX requests of
existing applications. End users only interact with the structured
spreadsheet, providing a predictable experience with clear affordances
for which customizations are possible.

Wildcard is currently an early prototype, with incomplete features and
limited coverage of sites. We plan to continue building site adapters
and testing the system with our own use cases, to better understand how
well the spreadsheet abstraction maps to real websites and customization
needs. Eventually we also plan to release the tool publicly, to
determine how end users choose to use spreadsheet-driven customization,
what usability challenges emerge, and how feasible it is for programmers
to build and maintain site adapters.

@@

To get a sense of how it might feel to use Wildcard, let's see an
example of someone using it to help with booking a trip on the travel
sites Airbnb and Expedia.{ These demos are best viewed as videos in the
online version of this paper
(\url{https://www.geoffreylitt.com/wildcard}).}

@@

The user starts by opening up the Airbnb search listings page to look
for a place to stay. The page looks nice and mostly works well, but is
missing some key features. As mentioned before, this page doesn't allow
the user to sort by price. It also doesn't let them filter by user
rating. Using Wildcard, the user can add these small features, while
leaving the page's design and the rest of its functionality unchanged.{
fig.~\ref{fig:airbnb-demo} shows an overview of augmenting the Airbnb
site.}

First, the user opens up the Wildcard panel, which shows a table
corresponding to the search results in the page. As they click around in
the table, the corresponding row in the page is highlighted so they can
see the connection between the views.

Then, the user can use standard spreadsheet column header features to
sort the page by price and filter by rating{
(fig.~\ref{fig:airbnb-demo}B)}.

After manipulating the data, the user was able to close the table view
and continue using the website with its original visual design. The
table view offers a way to change the data backing a page, but does not
need to replace the original interface entirely.

Most websites that show lists of data also offer actions that can be
taken on a row in the table, like adding an item to a shopping cart.
Wildcard has the ability to make these ``row actions'' available in the
data table if the site adapter implements them. The main advantage this
provides is the ability to easily perform a row action in bulk across
multiple rows.

For example, it's tedious on Airbnb to click on listings one by one to
add them to a list of favorites. Using Wildcard, the user can just
select multiple rows and favorite all of them with one click{
(fig.~\ref{fig:airbnb-demo}D)}. Similarly, we can also open the detailed
pages for many listings in new tabs.

Now the user would like to jot down some notes on the pros and cons of
each listing. To do this, they can simply type notes into an additional
column next to each listing, and the notes appear inside the listings in
the original UI{ (fig.~\ref{fig:airbnb-demo}A)}. These annotations are
also persisted in the browser for future retrieval.

Wildcard also includes a formula language which enables more
sophisticated tweaks that fetch external data and perform computations.

When traveling without a car, it's nice to evaluate potential places to
stay based on how walkable the surroundings. Using Wildcard formulas, we
can integrate Airbnb with Walkscore, an API that rates the walkability
of any location on a 1-100 scale. When we call the \texttt{WALKSCORE}
formula with the latitude and longitude of the listing, it returns the
walk score as the cell value. Because the cell's contents are injected
into the page, the score also shows up in the page body{
(fig.~\ref{fig:airbnb-demo}C)}.

@@

In previous examples the data extracted from the site was marked as
read-only; users cannot change the name or price of an Airbnb listing.
In this next case, the cells are marked as writable, so that changes in
the table are reflected in the form inputs. This becomes useful when
combined with GUI widgets for editing the value of a table cell.

Filling in dates for a flight search typically requires opening up a
separate calendar app to find the right dates, and then manually copying
them into the form. In Wildcard, we can make this easier by providing a
datepicker widget that has privileged access to the user's calendar
information{ (fig.~\ref{fig:expedia-demo}B)}.

Here we've presented just a few possibilities for how to use Wildcard.
We think the interactive data table offers a flexible computational
model that can support a wide range of other useful modifications, all
while remaining familiar and easy to use.

@@

In order to maximize extensibility, Wildcard is implemented as a small
core program along with several types of plugins: site adapters,
formulas, and cell renderers/editors. The core contains functionality
for displaying the data table and handling user interactions, and the
table implementation is built using the
\href{https://handsontable.com/}{Handsontable} Javascript library.

@@

Site adapters specify the bidirectional connection between the web page
and its structured data representation.

For extracting data from the page and getting it into structured form,
Wildcard provides ways to concisely express web scraping logic. For
example, here is a code snippet for extracting the name of an Airbnb
listing:

@@

Sometimes important data is not shown in the UI, making it impossible to
scrape from the DOM. To address this, we have also prototyped mechanisms
for site adapters to observe AJAX requests made by the browser and
extract data directly from JSON responses. This mechanism was used to
implement the Airbnb Walkscore example, since latitude and longitude
aren't shown in the Airbnb UI, but they are available in AJAX responses.
We also plan to add the ability for site adapters to scrape data across
multiple pages in paginated lists of results (as explored in prior work
\autocite{huynh2006}).

The site adapter also needs to support the reverse direction: sending
updates from the table to the original page. Most DOM manipulation is
not performed directly by the site adapter; instead, the adapter
specifies how to find the spans representing data rows, and the core
platform mutates the DOM to reflect the table state. The only exception
is row actions (like favoriting an Airbnb listing), which are
implemented as imperative Javascript functions that can can mutate the
DOM, simulate clicks on buttons, etc.

@@

The idea of spreadsheet-driven customization is guided by several design
principles, inspired by prior work and our own experimentation. We think
these principles can also broadly inform the design of end user
programming tools, especially those that enable users to customize
existing software.

@@

Today, most personal computing consists of using applications, which
bundle together behavior and data to provide some set of functionality.
While there are limited points of interoperability, applications
generally are designed to operate independently of one another. Once a
user has gotten past learning the basic idioms of modern computing like
windows and scrollbars, most of the effort invested in learning to use
one application does not carry over to using other applications.

@@

Spreadsheet-driven customization aims to port this UNIX philosophy to
the world of isolated Web applications, by creating a consistent data
structure to represent the data inside many applications. In UNIX, the
universal format is a text stream; in Wildcard, it is a relational
table: a simple abstraction generic enough to describe the data used in
many different applications. Because Wildcard maps the data from all
applications to the table format, users can invest in mastering the
Wildcard table editor, the formula language, and cell editor UIs, and
reuse those same tools to customize many different applications.

@@

Seymour Papert advocated for programming systems with a ``low floor,''
making it easy for novices to get started, as well as a ``high
ceiling,'' providing a large range of possibilities for more
sophisticated users \autocite{resnick2016}. Our goal is for
spreadsheet-driven customization to meet both of these criteria.
Although we need more testing with real users to know for sure, we think
Wildcard clearly provides a low floor, and likely has a flexible enough
foundation to eventually provide a high ceiling.

One of the most interesting properties of spreadsheets is the amount of
value they can provide to users who are aware of only a tiny sliver of
their functionality. Not only can a novice perform some basic activities
in a spreadsheet with almost no training (e.g., storing tables of
numbers or computing simple sums), these activities are often actually
valuable for the user! The fact that useful tasks can be performed early
on supports the user's natural motivation to continue using the tool,
and to eventually learn its more powerful features if needed
\autocite{nardi1991}. In contrast, many traditional programming systems
require an enormous upfront investment of time and practice before
someone is able to write a program that actually helps them achieve a
real task.

As part of ensuring a low floor for spreadsheet-driven customization, we
have focused on including genuinely valuable features for novices. For
example, a user can sort a table with a single click, or simply type in
some annotations. We would expect many Wildcard users to start out using
these simpler features before potentially moving on to more
sophisticated features like formulas.

Another aspect of providing a low floor is providing an ``in-place
toolchain'' \autocite{inkandswitch2019}---minimizing the effort of
moving from using to customizing, by making customization tools
available in the same environment where the user is already using the
software. This quality is distinct from the level of technical skill
needed to use the tool: for example, setting up a workflow trigger in an
end user programming system like \href{https://ifttt.com/}{IFTTT} does
not require much technical skill, but does require leaving the user's
normal software and entering a separate environment; conversely, running
a Javascript snippet in the browser console requires programming skills,
but can be done immediately and casually in the flow of using a website.

@@

Wildcard provides an in-place toolchain---on any site that supports
Wildcard, the user is one click away from starting to tweak the site
with minimal friction. There is no need for them to switch into another
environment or figure out how to extract data from the page. Once the
user starts editing, Wildcard also provides live feedback to help users
understand the changes they are making. Even if a user isn't yet totally
familiar with Wildcard, they can learn to use the system by trying
things out and seeing what happens.

Since we have still only built several site adapters and demos, it's
still too early to tell how high the ceiling is for the customizations
that can be achieved with Wildcard, but we think that with enough
operators the formula language could eventually support a wide variety
of customizations. We plan to explore this aspect further by trying to
solve more real problems with the system and observing where limitations
emerge in practice.

@@

Real-world spreadsheet usage in offices is highly collaborative: most
users just perform simple changes, while a few coworkers help with
writing more complex formulas or even programming macros
\autocite{nardi1990}. Inspired by this, we aim to make
spreadsheet-driven customization a collaborative activity that combines
the different abilities of many users in a collaborative ecosystem.

The main way we do this is by separating website customization into two
separate stages: structured data extraction, and using the resulting
spreadsheet. The first stage is currently only available to programmers
who can code site adapters in Javascript, whereas the second stage is
available for any non-programmer end user. This architecture frees end
users from needing to think about data extraction, and enables a
community of end users to reuse the efforts of programmers building site
adapters.

The group of users building adapters does not necessarily need to be
limited only to programmers. In the future we might explore enabling end
users to also create site adapters, drawing on related work on enabling
end users to extract structured data from websites
\autocite{chasins2018,huynh2006}. But even in that case, there would
still be a separation between motivated, tech-savvy end users building
adapters and more casual end users just using the spreadsheet view.

Another stakeholder to consider is the first party developers of the
original software. Spreadsheet-driven customization does not depend on
cooperation from first-party website developers: a third party
programmer can write an adapter for any website, which can access any
information available in the browser. On the other hand, if first
parties were to expose structured data from their applications, it would
avoid the need for adapters and generally make customization a lot
easier. We think there are compelling reasons for first parties to
consider doing this. Providing Wildcard support would allow users to
build extensions to fulfill their own feature requests. It also would
not necessarily require much effort: adding Wildcard support would be
more straightforward for a first-party than a third-party because they
have direct access to the structured data in the page. There is also
precedent for first parties implementing an official client extension
API in response to user demand: for several years, Google maintained an
official extension API in Gmail for Greasemonkey scripts to
use.\footnote{Incidentally, since then, third parties have continued to
  maintain stable Gmail extension APIs used by many browser extensions
  \autocite{streak,talwar2019}, illustrating the potential of
  collaboratively maintaining third party adapters.}

@@

While similar in high-level goals, Wildcard employs a different solution
strategy from these projects: whereas they generally require building
new software from scratch for that environment, Wildcard instead aims to
maximize the malleability of already existing software. This approach
has the pragmatic benefit of being immediately useful in more
situations, although it also requires working within more rigid
constraints.

With substantial future work, Wildcard could become more similar to
these other projects, growing from a platform for tweaking existing
software into a platform for building new software from scratch. This
would likely end up resembling existing tools for building
spreadsheet-driven applications (discussed more below), but with an
extra focus on customizability by end users of the software.

@@

Wildcard's approach is most similar to a class of tools that identify
structured data in a web page, and use that structure to support end
user modification of the page.

Sifter \autocite{huynh2006} enables users to sort and filter lists of
data on web pages, providing a result similar to Wildcard's sort and
filter functionality. The underlying mechanism is also similar: Sifter
extracts structured data from the page to enable its user-facing
functionality. Wildcard aims to extend this approach to support much
broader functionality than just sorting and filtering. In support of
this goal, Wildcard also shows the structured data table directly to the
user, whereas Sifter only shows sort and filter controls, without
revealing the underlying data table. The extraction mechanism is also
different: Sifter uses a combination of automated heuristics and
interactive user feedback to extract data, whereas Wildcard currently
relies on programmers creating wrappers for extracting structured data,
likely leading to higher quality extraction but on fewer sites.

Thresher \autocite{hogue2005} enables users to create wrappers which map
unstructured website content to Semantic Web content. Like Wildcard and
Sifter, Thresher augments the experience of original page based on
identifying structure: once semantic content has been identified, it
creates context menus in the original website which allow users to take
actions based on that content. Wildcard and Thresher share an overall
approach but focus on complementary parts of the problem: Thresher aims
to enable end users to create content wrappers, but the actions
available on the structured data are created by programmers; conversely,
Wildcard delegates wrapper creation to programmers but gives end users
more flexibility to use the structured data in an open-ended way.

@@

``Sloppy programming'' \autocite{little2010} tools like Chickenfoot
\autocite{bolin2005} and Coscripter \autocite{leshed2008} enable users
to create scripts that perform actions like filling in text boxes and
clicking buttons, without directly interacting with the DOM. Users
express the desired page elements in natural, informal terms
(e.g.~writing ``the username box'' to represent the textbook closest to
the label ``username''), and then using heuristics to determine which
elements most likely match the user's intent. This approach allows for
expressing a wide variety of commands with minimal training, but it also
has downsides. It is difficult to know whether a command will
consistently work over time (in addition to changes to the website,
changes to the heuristics can also cause problems), and it is not easy
for users to discover the space of possible commands
\autocite{little2010}.

Wildcard offers a sharp contrast to sloppy programming, instead choosing
to expose a high degree of structure through the familiar spreadsheet
table. Wildcard offers more consistency: for example, clicking a sort
header will always work correctly as long as the site adapter is
maintained. Wildcard also offers clearer affordances for what types of
actions are possible, or, crucially, what actions are \emph{not}
possible, which is useful to know. On the other hand, Wildcard cannot
offer coverage of all websites, and has a narrower range of possible
actions than sloppy tools. We expect that with enough site adapters and
formulas, these downsides can be mitigated.

@@

Prior work has made the powerful realization that a spreadsheet can
serve as an end-user-friendly backing data store and computation layer
for an interactive web application. Research projects like Object
Spreadsheets \autocite{mccutchen2016}, Quilt \autocite{benson2014},
Gneiss \autocite{chang2014}, and Marmite \autocite{wong2007}, as well as
commercial tools like Airtable Blocks \autocite{zotero-79} and Glide
\autocite{zotero-81} allow users to view data in a spreadsheet table,
compute over the data using formulas, and connect the table to a GUI.
Because many users are already familiar with using spreadsheets, this
way of creating applications tends to be far easier than traditional
software methods; for example, in a user study of Quilt, many people
were able to create applications in under 10 minutes, even if they
expected it would take them many hours.

@@

Web scraping tools focus on extracting structured data out of
unstructured web pages. Web scraping is closely related to the
implementation of Wildcard, but has different end goals: web scraping
generally extracts static data for processing in another environment,
whereas Wildcard modifies the original page by maintaining a
bidirectional connection between the extracted data and the page.

@@

There are still many open questions about spreadsheet-driven
customization, which we hope to answer through targeted development and
usage of the Wildcard prototype.

The most important question is whether the computational model can
support a large variety of useful customizations. While initial demos
are promising, we need to develop more site adapters and use cases to
explore this further. We plan to continue privately testing the system
with our own needs, and to eventually deploy the tool publicly, once the
API is stable enough and can support a critical mass of sites and use
cases. We also plan to run usability studies to evaluate and improve the
design of the tool.

We suspect that two areas of the current model may prove too limiting.
First, Wildcard's data model only shows a single table at a time,
without any notion of relationships between tables. A richer data model
with foreign keys might be necessary to support certain use cases. For
designing a tabular interface on top of a richer data model, we could
learn from the interface of Airtable which shows related objects in a
table cell, or add nested rows as used in other systems
\autocite{mccutchen2016,bakke2016}.

Second, there is no mechanism for end users to express imperative
workflows (e.g.~``click this button, and then\ldots{}''); they can only
write formulas that return data and then inject the resulting data into
the page. If it proves useful, we might consider adding a system for
expressing workflows like this, although it would substantially alter
the computational model and it's not clear how it would fit together
with the existing table view.

>>>>

Web applications often don't match the particular needs of their users.
Sometimes there is a browser extension available to patch an issue, and
if the user is a programmer they might be able to fix it themselves. But
for most people, the only recourse is to complain to the developers, or
more likely, to simply give up. Back in 1977, in \emph{Personal Dynamic
Media} \autocite{kay1977}, Alan Kay envisioned personal computing as a
medium that let a user ``mold and channel its power to his own needs,''
but today, software feels more like concrete than clay.

In this paper, we present \emph{spreadsheet-driven customization}, a
technique that enables end users to customize software. The idea is to
augment an application's UI with a spreadsheet that is synchronized with
the application's data. When the user manipulates the spreadsheet, the
underlying data is modified and the changes are propagated to the UI,
and vice versa. We have implemented this technique in a prototype
browser extension called Wildcard. Section 2 presents demos of using
Wildcard to augment real websites in useful ways, and Section 3
describes the implementation of the extension.

Spreadsheet-driven customization provides an easy entry point for end
users, since small tweaks can be performed with a single click. At the
same time, it also supports a variety of richer customizations, like
adding private annotations to a webpage or joining in related data from
a web API. In Section 4, we elaborate on this principle of ``low floor,
high ceiling,'' as well as other design principles guiding our work.

Prior work \autocite{mccutchen2016,benson2014,chang2014} has enabled end
users to create ``spreadsheet-driven applications'' which use
spreadsheets as a backing data layer. Spreadsheet-driven
\emph{customization} applies this idea in a different context:
customizing existing software, rather than building new software from
scratch. Our technique does not require that the application actually be
backed by a spreadsheet; it merely uses the spreadsheet as an interface
for viewing and modifying the internal state of the application.

This approach requires extracting structured data from the user
interfaces of existing applications, but we hide the complexity of data
extraction from end users. Programmers write \emph{site adapters} which
use web scraping techniques to extract structured data from existing
applications and map them to the spreadsheet table. End users only
interact with the structured spreadsheet, providing a straightforward
customization experience. In Section 5, we explain this architecture in
more depth, and describe how Wildcard relates to existing work on
malleable software, web customization, spreadsheet-driven applications,
and web scraping.

The Wildcard extension is currently an early research prototype. We plan
to continue testing the system with our own use cases to understand
better how well the spreadsheet abstraction maps to real websites and
customization needs. Eventually we plan to release the tool publicly, in
order to study how end users choose to customize applications, discover
usability challenges, and to test the feasibility of programmers
building and maintaining site adapters.

@@

To get a sense of the user experience of using Wildcard, let's see an
example of someone using it to help with booking a trip.{ These demos
are best viewed as videos in the online version of this paper
(\url{https://www.geoffreylitt.com/wildcard}).}

@@

In 2012, the travel site Airbnb removed the ability to sort
accommodation listings by price. Users could still filter by price
range, but could no longer view the cheapest listings first. Many users
complained on online message boards that the change seemed hostile to
users. ``It's so frustrating!..What is the logic behind not having this
function?'' said one user on the
\href{https://community.withairbnb.com/t5/Hosting/Sorting-listing-by-price/td-p/559404}{Airbnb
support forum}. Alas, the feature remains missing to this day.

Using Wildcard, the user can fix this omission, while leaving the page's
design and the rest of its functionality unchanged.{
fig.~\ref{fig:airbnb-demo} shows an overview of augmenting the Airbnb
site.} First, the user opens up the Wildcard panel, which shows a table
corresponding to the search results in the page. As they click around in
the table, the corresponding row in the page is highlighted so they can
see the connection between the views.

Then, the user clicks on the price column header to sort the
spreadsheet, and the Airbnb UI, by in ascending order by price{
(fig.~\ref{fig:airbnb-demo}B)}. They can similarly use the column
headers to filter to listings with a user rating above 4.5, which the
original Airbnb UI also doesn't allow.

After manipulating the data, the user closes the table view and
continues using the website with its original visual design. The table
view offers a way to change the data backing a page, but does not need
to replace the original interface entirely.

Most websites that show lists of data also offer actions on rows in the
table, like adding an item to a shopping cart. Wildcard has the ability
to make these ``row actions'' available in the data table through the
site adapter.

In the Airbnb UI, saving multiple listings to a Favorites list requires
tediously clicking through them one by one. Using Wildcard, the user can
select multiple rows and favorite all of them with a single click{
(fig.~\ref{fig:airbnb-demo}D)}. Similarly, the user can also open the
detailed pages for many listings at once.

Now the user wants to jot down some notes on the pros and cons of each
listing. To do this, they type some notes into an additional column in
each listing row, and the notes appear inside the listings in the
original UI{ (fig.~\ref{fig:airbnb-demo}A)}. The annotations are saved
in the browser for future sessions.

Wildcard also includes a formula language, which enables more
sophisticated customizations that fetch external data and perform
computations.

When traveling without a car, it's useful to evaluate potential places
to stay based on how walkable the surroundings are. Using Wildcard
formulas, the user can integrate Airbnb with Walkscore, an API that
rates the walkability of any location on a 1-100 scale. When they type
in a \texttt{WALKSCORE} formula with the latitude and longitude of the
listing, it returns the walk score as the cell's value. Because the
cell's contents are injected into the page, the score also shows up in
the page body{ (fig.~\ref{fig:airbnb-demo}C)}.

@@

In previous examples the table cells were read-only; users cannot change
the name or price of an Airbnb listing. In this next case, the cells are
writable, so that changes in the table are reflected in the form inputs.
This becomes especially useful when combined with GUI widgets that can
edit the value of a table cell.

Filling in dates for a flight search typically requires opening up a
separate calendar app to find the right dates and then manually copying
them into the form. In Wildcard, the user can avoid this cumbersome
workflow by using a datepicker widget that includes the user's personal
calendar information{ (fig.~\ref{fig:expedia-demo}B)}. The user can
directly click on the right date, and it gets inserted into the
spreadsheet and the original form.

Here we've presented just a few use cases for spreadsheet-driven
customization, but we think the interactive data table is flexible
enough to support a wide range of other useful modifications, all while
remaining familiar and easy to use.

@@

In order to promote extensibility, Wildcard is implemented as a small
core program along with several types of plugins: site adapters,
formulas, and cell renderers/editors. The core contains functionality
for displaying the data table and handling user interactions, and the
table implementation is built using the
\href{https://handsontable.com/}{Handsontable} Javascript library.

@@

Site adapters are a key part of Wildcard, since they specify the
bidirectional connection between the web page and its structured data
representation.

Wildcard provides an interface for concisely expressing how the DOM
should be mapped to a data table. For example, here is a code snippet
for extracting the name of an Airbnb listing:

@@

Sometimes important data is not shown in the UI, making it impossible to
scrape from the DOM. To address this, we have also prototyped mechanisms
for site adapters to observe AJAX requests made by the browser and
extract data directly from JSON responses. This mechanism was used to
implement the Airbnb Walkscore example, since latitude and longitude
aren't shown in the Airbnb UI, but they are available in AJAX responses.
In the future we also plan to add the ability for site adapters to
scrape data across multiple pages in paginated lists of results (as
explored in prior work \autocite{huynh2006}).

The site adapter also needs to support the reverse direction: sending
updates from the table to the original page. Most DOM manipulation is
not performed directly by the site adapter. Instead, Wildcard
automatically mutates the DOM to reflect the table state, using the same
declarative spec shown above. The only exception is row actions (like
favoriting an Airbnb listing), which are implemented as imperative
Javascript functions that can can mutate the DOM, simulate clicks on
buttons, etc.

@@

The idea of spreadsheet-driven customization is guided by several design
principles, inspired by prior work and our own experimentation. We think
these principles can also broadly inform the design of tools for end
user software customization.

@@

Today, most personal computing consists of using applications, which
bundle together behavior and data to provide some set of functionality.
While there are some limited points of interoperability, applications
generally are designed to operate independently of one another.

@@

Spreadsheet-driven customization aims to bring some of this UNIX ethos
to the world of isolated Web applications, by creating a consistent data
structure to represent the data inside many applications. In UNIX, the
universal format is a text stream; in Wildcard, it is a relational
table. The table is a simple abstraction which is also generic enough to
describe the data used in many different applications. Because Wildcard
maps the data from all applications to the table format, users can
invest in mastering the Wildcard table editor, the formula language, and
cell editor UIs, and reuse those same tools to customize many different
applications.

@@

Seymour Papert advocated for programming systems to have a ``low
floor,'' making it easy for novices to get started, and a ``high
ceiling,'' providing a large range of possibilities for more
sophisticated users \autocite{resnick2016}. Our goal is for
spreadsheet-driven customization to meet both of these criteria.

One of the most interesting properties of spreadsheets is that users who
are only know a tiny sliver of their functionality (e.g., storing tables
of numbers or computing simple sums) can still use them in genuinely
valuable ways. The fact that useful tasks can be performed early on
supports the user's natural motivation to continue using the tool, and
to eventually learn its more powerful features if needed
\autocite{nardi1991}. In contrast, many traditional programming systems
require an enormous upfront investment of time and practice before
someone is able to write a program that helps them achieve a useful task
in their life.

As part of ensuring a low floor, we have focused on including genuinely
valuable features for novices. For example, a user can sort a table with
a single click, or simply type in some annotations. We would expect many
Wildcard users to start by using these simpler features before
potentially moving on to more sophisticated features like formulas.

Another aspect of providing a low floor is providing an ``in-place
toolchain'' \autocite{inkandswitch2019}: minimizing the effort of moving
from using to customizing, by making customization tools available in
the same environment where the user is already using the software. This
quality is distinct from the level of technical skill needed to use the
tool. For example, setting up a workflow trigger in an end user
programming system like \href{https://ifttt.com/}{IFTTT} does not
require much technical skill, but does require leaving the user's normal
software and entering a separate environment; conversely, running a
Javascript snippet in the browser console requires programming skills,
but can be done immediately and casually in the flow of using a website.

@@

Wildcard provides an in-place toolchain because the spreadsheet can be
instantly opened in the browser window while using any supported
website. Once the user starts editing, Wildcard also provides live
feedback, so that even if a user isn't yet totally familiar with
Wildcard, they can learn to use the system through experimentation.

Since we have still only built several site adapters and demos, it is
still too early to tell exactly how high the ceiling is for the
customizations that can be achieved with Wildcard. But we think that
with enough operators, the formula language could support a wide variety
of customizations---people have managed to use simple spreadsheet
formula languages to solve a surprisingly large range of problems. We
plan to explore this aspect further by trying to solve more real
problems with the system and observing where limitations emerge in
practice.

@@

Real-world spreadsheet usage in offices is highly collaborative: most
users just perform simple changes, while a few coworkers help with
writing more complex formulas or even programming macros
\autocite{nardi1990}. Inspired by this collaborative approach, we aim to
make spreadsheet-driven customization a collaborative activity that
combines the different abilities of many users.

The main way we do this is by separating website customization into two
separate stages: structured data extraction, and using the resulting
spreadsheet. The data extraction stage is currently only available to
programmers who can code site adapters in Javascript, whereas the
customization stage is available to any non-programmer end user. This
architecture frees end users from needing to think about data
extraction, and enables a community of end users to reuse the efforts of
programmers building site adapters.

The group of users building adapters does not necessarily need to be
limited only to programmers. In the future, we plan to explore enabling
end users to also create site adapters, drawing on related work in this
area \autocite{chasins2018,huynh2006}. But even in that case, we still
envision a separation between highly motivated, tech-savvy end users
building adapters, and more casual end users just using the spreadsheet
view.

Another stakeholder to consider is the first party developers of the
original software. Spreadsheet-driven customization does not depend on
cooperation from first-party website developers, but if they were to
expose structured data in their web application clients, it would make
site adapters unnecessary, producing more robust results with much less
work.

We think there are compelling reasons for first parties to consider
doing this. Providing Wildcard support would allow users to build
extensions to fulfill their own feature requests. It also would not
necessarily require much effort: adding Wildcard support could be made
fairly straightforward for a first-party that already has direct access
to the structured data in the page. There is also precedent for first
parties implementing an official client extension API in response to
user demand: for several years, Google maintained an official extension
API in Gmail for Greasemonkey scripts to use.\footnote{Incidentally,
  since then, third parties have continued to maintain stable Gmail
  extension APIs used by many browser extensions
  \autocite{streak,talwar2019}, illustrating the potential of
  collaboratively maintaining third party adapters.}

@@

While similar in broad goals, Wildcard employs a different solution
strategy. These projects generally require building software from
scratch in a new environment, whereas Wildcard aims to maximize the
malleability of software built with existing tools. This approach has
the pragmatic benefit of enabling customization for much more software,
although it also imposes more rigid constraints.

With substantial future work, we think Wildcard could become more
similar to these other projects, growing from a platform for tweaking
existing software into a platform for building new software from
scratch. This would likely end up resembling existing tools for building
spreadsheet-driven applications (discussed more below), but with an
additional focus on customizability by end users of the software.

@@

Wildcard's solution approach is most similar to other tools that
identify structured data in a web page, and then use that structure to
support end user customization of the page.

Sifter \autocite{huynh2006} enables users to sort and filter lists of
data on web pages, which resembles Wildcard's sort and filter
functionality. The underlying mechanism is also similar, since Sifter
extracts structured data from the page to enable its user-facing
functionality. The systems also have major differences. Wildcard aims to
extend this approach to support much broader functionality, and shows
the structured data table directly to the user. In contrast, Sifter only
supports sorting and filtering, and does not reveal the underlying data
table. The extraction mechanism is also different: Sifter uses a
combination of automated heuristics and interactive user feedback to
enable end users to extract data, whereas Wildcard hides data extraction
from end users.

Thresher \autocite{hogue2005} enables users to create wrappers which map
unstructured website content to Semantic Web content. Like Wildcard and
Sifter, Thresher augments the experience of original page based on
identifying structure; once semantic content has been identified, it
creates context menus in the original website which allow users to take
actions based on that content. Wildcard and Thresher focus on
complementary parts of the problem. Thresher aims to enable end users to
create content wrappers, but the actions available on the structured
data are created by programmers; conversely, Wildcard delegates wrapper
creation to programmers but gives end users more flexibility to use the
structured data in an open-ended way.

@@

``Sloppy programming'' \autocite{little2010} tools like Chickenfoot
\autocite{bolin2005} and Coscripter \autocite{leshed2008} enable users
to create scripts that perform actions like filling in text boxes and
clicking buttons, without directly interacting with the DOM. Users
express the desired page elements in natural, informal terms
(e.g.~writing ``the username box'' to represent the textbook closest to
the label ``username''), and then using heuristics to determine which
elements most likely match the user's intent. This approach allows for
expressing a wide variety of commands with minimal training, but it also
has downsides \autocite{little2010}:

\begin{itemize}
\tightlist
\item
  Reliability: It can be difficult to know whether a command will
  consistently work over time. In addition to changes to the website,
  changes to the heuristics can also cause problems.
\item
  Discoverability: it can be difficult for users to discover the space
  of possible commands.
\end{itemize}

Wildcard offers a sharp contrast to sloppy programming, instead choosing
to expose a high degree of structure through the familiar spreadsheet
table. Wildcard offers more consistency; for example, clicking a sort
header will always work correctly as long as the site adapter is
maintained. Wildcard also offers clearer affordances for what types of
actions are possible, or, crucially, what actions are \emph{not}
possible, which is useful to know. On the other hand, Wildcard's
explicit site adapter approach will result in coverage of many fewer
websites, and also a narrower range of possible actions than ``sloppy''
tools.

@@

Many others have made the powerful realization that a spreadsheet can
serve as an backing data store and computation layer for an interactive
web application, enabling end users to create such applications more
easily. Research projects like Object Spreadsheets
\autocite{mccutchen2016}, Quilt \autocite{benson2014}, Gneiss
\autocite{chang2014}, and Marmite \autocite{wong2007}, as well as
commercial tools like Airtable Blocks \autocite{zotero-79} and Glide
\autocite{zotero-81} allow users to view data in a spreadsheet table,
compute over the data using formulas, and then connect the table to a
GUI. Because many users are already familiar with using spreadsheets,
this way of creating applications tends to be much easier than
traditional software methods; for example, in a user study of Quilt,
many users were able to create applications in under 10 minutes, even if
they expected it would take them many hours.

@@

Web scraping tools focus on extracting structured data out of
unstructured web pages. Web scraping is closely related to the
implementation of Wildcard, but has different end goals: web scraping
generally extracts static data for processing in another environment,
whereas Wildcard customizes the application's UI by maintaining a
bidirectional connection between the extracted data and the page.

@@

There are still many open questions about spreadsheet-driven
customization which we hope to answer through targeted development and
usage of the Wildcard prototype.

The most important question is: where are the limits of this
computational model? What types of useful customizations can it support
or not support? While initial demos suggest a variety of use cases, we
plan to develop more site adapters and demos to explore this question
further. We will start with privately testing the system with our own
needs, and then eventually deploy the tool publicly, once the API is
stable enough and can support a critical mass of sites and use cases. We
also plan to run usability studies to evaluate and improve the design of
the tool.

We suspect that two areas of the current model may prove limiting.
First, Wildcard's data model currently shows a single table at a time,
without any notion of relationships between tables. A richer data model
with foreign keys might help support certain use cases. For designing a
spreadsheet interface on top of a richer relational model, we could
learn from other systems with this design
\autocite{mccutchen2016,bakke2016}.

A second potential limitation is that there is currently no mechanism
for end users to express imperative workflows with sequences of actions,
which related work has shown to be useful
\autocite{bolin2005,leshed2008}. It's not clear how such workflows could
fit into Wildcard, since spreadsheets have a fundamentally different
computation model, and site adapters created by programmers cannot
easily account for every possible action a user would want to take in a
page.
